---
title: "chromatographR: An introduction to HPLC-DAD analysis"
author: "Ethan Bass"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{chromatographR: An introduction to HPLC-DAD analysis}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
library("knitr", quietly=TRUE)
library(chromatographR)
library(parallel)
opts_chunk$set(prompt = TRUE, highlight = FALSE, comment = NA, 
               dpi = 200, fig.width = 5, out.width = "600px")
suppressMessages(require(chromatographR, quiet=TRUE))
```

## Introduction

*chromatographR* is a package for the reproducible analysis of HPLC-DAD data in `R`. Liquid chromatography coupled to diode-array detection (HPLC-DAD) remains one of the most popular analytical methodologies due to its convenience and low-cost. However, there are currently very few open-source tools available for analyzing HPLC-DAD chromatograms or other "simple" chromatographic data. The use of proprietary software for the analysis of HPLC-DAD data is currently a significant barrier to reproducible science, since these tools are not widely accessible, and usually require users to select complicated options through a graphical user interface which cannot easily be repeated. Reproducibility is much higher in command line workflows, like *chromatographR*, where the entire analysis can be stored and easily repeated by anyone using publicly available software.

The *chromatographR* package began as a fork from the previously published *alsace* package [@wehrens2015], but has been reworked with improved functions for peak-finding, integration and peak table generation as well as a number of new tools for data visualization and downstream analysis. Unlike *alsace*, which emphasized multivariate curve resolution through alternating least squares (MCR-ALS), *chromatographR* is developed around a more conventional workflow that should seem more familiar to users of standard software tools for HPLC-DAD analysis. *chromatographR* includes tools for a) pre-processing, b) retention-time alignment, c) peak-finding, d) peak-integration and e) peak-table construction, as well as additional functions useful for analyzing the resulting peak table.

## Workflow

### Loading data

chromatographR can import data from a growing list of proprietary file formats using the `read_chroms` function from the chromConverter package. Supported file formats include 'Agilent ChemStation' and 'MassHunter' (`.D`) files, 'Thermo Raw' (`.raw`), 'Chromeleon' UV ASCII (`.txt`), 'Waters ARW' (`.arw`), 'Shimadzu' ASCII (`.txt`), and more. (For a complete list, see the [chromConverter documentation](https://ethanbass.github.io/chromConverter/)). Select the appropriate file format by specifying the `format_in` argument (e.g., `csv`, `chemstation_uv`, `masshunter_dad`, `chromeleon_uv`, `waters_arw`, etc).

```{r load-data, eval = FALSE}
# single folder
read_chroms(paths = path, format_in = "chemstation_uv")

# multiple folders
path <- 'foo' # path to parent directory
folders <- list.files(path = path, full.names = TRUE)
dat <- read_chroms(folders, format_in = "chemstation_uv")
```

### Example data

The package includes some example data consisting of root extracts from tall goldenrod (*Solidago altissima*). Roots were extracted in 90% methanol and run on an Agilent 1100 HPLC coupled to a DAD detector, according to a previously described method [@uesugi2013]. The dataset is called `Sa` (abbreviated from *Solidago altissima*).

```{r load-example-data}
data(Sa)
```

### Pre-processing data

Data from liquid chromatography often suffer from a variety of non-informative artifacts, such as noise or a drifting baseline. In addition, the data produced by the instrument may have a higher resolution or wider range (along either the time or spectral dimensions) than we require. Fortunately, most of these issues can be remedied fairly easily. For example, smoothing can reduce noise in the spectral direction while baseline subtraction can help correct a drifting baseline. Interpolation of wavelengths and retention times can be used to reduce the dimensionality of the data, facilitating comparison across samples and reducing the time and computational load required by downstream analyses. All of these functions (smoothing, baseline correction, and interpolation) are available through the `preprocess` function and are enabled by default.

To select a narrower range of times and/or wavelengths, arguments can be provided to the optional `dim1` and `dim2` arguments. The `baseline_cor` function from the *ptw* package [@bloemberg2010] takes arguments `p` (an asymmetry parameter) and `lambda` (a smoothing parameter). You can read more about these in the documentation for `ptw::asysm`. You may want to experiment with these parameters before choosing values to use on your whole dataset.

```{r choose-parameters-for-fitting-baseline, fig.cap = "Comparison of baseline correction parameters: raw data (black), mild correction (green, p = 0.001), strong correction (red, p = 0.1)."}
i <- 2 # chromatogram number in list of data
tpoints <- get_times(Sa)
wavelength <- '200.00000' # this must match one of the column names from your chromatogram

matplot(x = tpoints, y = Sa[[i]][,wavelength],
       type = 'l', ylab = 'Abs (mAU)', xlab = 'Time (min)')
matplot(x = tpoints, y = ptw::baseline.corr(Sa[[i]][,wavelength], p = .001, lambda = 1e5),
       type = 'l', add = TRUE, col='darkgreen', lty = 3)
matplot(x = tpoints, y = ptw::baseline.corr(Sa[[i]][,wavelength], p = .1, lambda = 1e5),
       type = 'l', add = TRUE, col='firebrick2', lty = 3)
```

After selecting parameters for baseline correction, you can proceed with the pre-processing step as shown below. For 2D data (e.g., from a single-wavelength detector), the `dim2` argument should be left blank.

```{r preprocessing}
# choose dimensions for interpolation
new_times <- seq(10, 18.66, by = .01) # choose time-points
new_wavelengths <- seq(200, 318, by = 2) # choose wavelengths

dat.pr <- preprocess(Sa, dim1 = new_times, dim2 = new_wavelengths, p = .001, 
                     lambda = 1e5, cl = 1)
```

### Alignment

In many cases, liquid chromatography can suffer from retention time shifts of varying degree (e.g., due to temperature fluctuations, column degradation, or subtle changes in mobile-phase composition), which can make it very difficult to compare peaks across samples. Luckily, a number of "time-warping" algorithms have been developed for correcting these kinds of shifts. In `chromatographR`, parametric time warping (`ptw`) [@eilers2004; @bloemberg2010] and variable penalty dynamic time warping (`vpdtw`) [@clifford2009; @clifford2012] are available for correcting retention time shifts through the `correct_rt` function. Both warping functions aim to produce a better alignment of features by "warping" the time-axis of each supplied chromatogram to match a reference chromatogram. (The reference chromatogram can either be determined algorithmically or selected manually by setting the `reference` argument).

First, we check the alignment of our chromatograms using the `plot_chroms` function.

```{r check-alignment, fig.cap = "Chromatographic traces of four *S. altissima* root chromatograms at 210 nm."}
plot_chroms(dat.pr, lambdas = 210, show_legend = TRUE)
```

Our chromatograms appear to be shifted, as shown by the poor overlap of the chromatograms across our four samples. To remedy this problem, we can try to "warp" our chromatograms using one of the two options described above.

#### Parametric time warping

The `ptw` option can take a single wavelength or a list of wavelengths provided by the user using the `lambdas` argument. For each chromatogram, `ptw` then produces a "global" warping function across all the wavelengths supplied by the user. The code block below creates warping models for the samples in the provided list of data matrices. The same function is then used to warp each chromatogram according to the corresponding model, by setting the `models` parameter. Depending on the variety of your samples and the severity of the retention time shifts, it may take some experimentation with the warping parameters to get satisfactory results. Sometimes less can actually be more here -- for example, wavelengths with fewer peaks may sometimes yield better warping models. (Also see the documentation for [ptw](https://CRAN.R-project.org/package=ptw) for more guidance on warp function optimization).

```{r create-warp-models}
warping.models <- correct_rt(dat.pr, alg = "ptw", what = "models", lambdas = 210, scale = TRUE)
warp_ptw <- correct_rt(chrom_list = dat.pr, models = warping.models, what = "corrected.values")
```

We can then use the following code snippet to compare the alignment of warped (top panel) and unwarped (bottom panel) chromatograms.

```{r alignment-plot, fig.height = 6, fig.cap = "Comparison of parametric time warping (PTW) aligned chromatograms and raw data."}
par(mfrow=c(2,1))
plot_chroms(dat.pr, lambdas = 210, show_legend = FALSE)
legend("topleft", legend = "Raw data", bty = "n")

plot_chroms(warp_ptw, lambdas = 210, show_legend = FALSE)
legend("topleft", legend = "PTW", bty = "n")
```

Clearly, the alignment is considerably improved after warping. You can also use the `correct_rt` function to do a global alignment on multiple wavelengths, by providing a list of wavelengths to the `lambdas` argument, but this will not always improve the results. While the alignment still isn't perfect after warping, it is probably good enough to align our peaks and assemble them in the peak table, which is our primary goal.

#### Variable penalty dynamic time warping

Variable penalty dynamic time warping is another algorithm that can be very effective for correcting retention time shifts [@clifford2009; @clifford2012]. Instead of calculating an explicit warping function, dynamic time warping (DTW) works by repeatedly shifting, expanding or contracting the time axis in small steps until a satisfactory alignment is attained. This process can result in very close matching of peaks, but can also lead to severe peak distortion if the time axis is repeatedly expanded and/or contracted [@tomasi2004]. **Variable penalty dynamic time warping** is a derivative of DTW that constrains expansion and contraction of the time-axis by applying a penalty to these types of "risky" moves. This can often achieve a very good alignment with minimal peak distortion. VPdtw package is a suggested dependency since it requires a working C++ compiler. Unlike `PTW`, `VPdtw` can only take a single wavelength as input (but it returns a global warping on the whole chromatographic matrix). The code snippet below uses VPdtw to warp the example chromatograms using the first chromatogram as a reference. The `penalty` and `maxshift` parameters can be adjusted to control the degree of warping.

```{r vpdtw-alignmment, eval = TRUE}
warp_vpdtw <- correct_rt(chrom_list = dat.pr, alg = "vpdtw", lambdas = 210, 
                   what = "corrected.values")
```

```{r plot-chroms, fig.height = 6, fig.cap = "Comparison of variable penalty dynamic time warping (VPdtw) aligned chromatograms (top) with raw data (bottom)."}
par(mfrow=c(2,1))
plot_chroms(dat.pr, lambdas = 210, show_legend = FALSE)
legend("topleft", legend = "Raw data", bty = "n")

plot_chroms(warp_vpdtw, lambdas = 210, show_legend = FALSE)
legend("topleft", legend = "VPdtw", bty = "n")
```

Chromatograms can also be visualized as heatmaps with `plot_chroms_heatmap`, which can be a powerful way to check the alignment of chromatograms across samples.

```{r plot-chroms-heatmap, fig.height = 6, fig.cap = "Comparison of variable penalty dynamic time warping (VPdtw) aligned chromatograms (top) with raw data (bottom) via heatmap plot."}
par(mfrow=c(3,1))
plot_chroms_heatmap(dat.pr, lambdas = 210, show_legend = FALSE,
                    title="Raw data")
plot_chroms_heatmap(warp_ptw, lambdas = 210, show_legend = FALSE,
                    title="Parametric Time Warping (PTW)")
plot_chroms_heatmap(warp_vpdtw, lambdas = 210, show_legend = FALSE,
                    title="Variable Penalty Dynamic Time Warping (VPdtw)")
```

In this case the alignments produced by the two algorithms are quite similar, but in other cases one algorithm may dramatically outperform the other. In my experience, alignments generated by VPdtw are less prone to overfitting. However, the PTW algorithm is more interpretable since it fits an explicit model to the data. Examination of the coefficients from these models can often be useful for the identification of problematic samples.

### Peak detection and integration

The `get_peaks` function produced a nested list of peaks by looping through the supplied chromatograms at the specified wavelengths, finding peaks, and fitting them to the specified function using non-linear least squares. The area under the curve for each peak is then estimated using trapezoidal approximation. The `fit` argument can be used to specify a peak-fitting model. By default, peaks will be fit to an exponential-gaussian hybrid function (`"egh"`) [@lan2001]. Other options include the bidirectional exponentially modified gaussian (`"bemg"`) or `"gaussian"`. Alternatively, peak areas can be integrated without applying a model (`fit = "raw"`). The function returns a nested list of data.frames containing parameters for the peaks identified in each chromatogram.

```{r get-peaks, message = FALSE, warning = FALSE}
# find and integrate peaks using exponential-gaussian hybrid model
pks_bemg <- get_peaks(warp_vpdtw, lambdas = c(210), sd_max = 40, fit = "bemg")

# find and integrate peaks using gaussian peak fitting
pks_gauss <- get_peaks(warp_vpdtw, lambdas = c(210), sd_max = 40, fit = "gaussian")

# find and integrate peaks using exponential-gaussian hybrid model
pks_egh <- get_peaks(warp_vpdtw, lambdas = c(210), sd_max = 40, fit = "egh")
```

#### Visualization

The `plot.peak_list` function allows you to visually assess the peak integration accuracy. Below we compare the peaks fitted for the first chromatogram (`idx = 1`) using the two algorithms. Usually the egh function performs slightly better for asymmetrical peaks. Even though the peaks are fully filled in when the `raw` setting is selected, the results may not necessarily be more accurate.

```{r plot-peaks, warning = FALSE, fig.height = 6, fig.cap = "Comparison of peak fitting algorithms: gaussian (top), exponential-gaussian hybrid (middle) and raw (bottom)."}
par(mfrow=c(3,1))
plot(pks_gauss, idx = 1, lambda = 210)
legend("topright", "Gaussian", bty = "n")

plot(pks_egh, idx = 1, lambda = 210)
legend("topright", "Exponential-gaussian hybrid", bty = "n")

plot(pks_bemg, idx = 1, lambda = 210)
legend("topright", "Bidirectional exponentially modified gaussian", bty = "n")
```

#### Local baseline correction

When peaks sit on a rising or falling background — for example, during gradient elution — the `baseline` argument can be used to fit a local baseline beneath each peak. Setting `baseline = "flat"` fits a constant offset per peak; `baseline = "sloped"` fits a linearly varying baseline across the peak window. In both cases peak heights and areas are computed above the fitted baseline.

The effect is most visible on data that has not been baseline-corrected during preprocessing. Below, we preprocess the raw data without baseline correction to better illustrate the difference.

```{r baseline-fitting, warning = FALSE, fig.height = 6, fig.cap = "Comparison of peak fitting with no baseline (top), a flat baseline (middle), and a sloped baseline (bottom) on un-baseline-corrected data. Shaded areas represent the fitted peak above the local baseline."}
dat.pr.ubc <- preprocess(Sa[2], dim1 = new_times, dim2 = new_wavelengths,
                        remove.time.baseline = FALSE, cl = 1)

pks_ubc <- get_peaks(dat.pr.ubc, lambdas = 210, fit = "egh", cl = 1)
pks_flat <- get_peaks(dat.pr.ubc, lambdas = 210, fit = "egh", baseline = "flat", cl = 1)
pks_sloped <- get_peaks(dat.pr.ubc, lambdas = 210, fit = "egh", baseline = "sloped", cl = 1)

par(mfrow=c(3,1))
plot(pks_ubc, lambda = 210)
legend("topright", "Uncorrected", bty = "n")

plot(pks_flat, lambda = 210)
legend("topright", "Flat baseline", bty = "n")

plot(pks_sloped, lambda = 210)
legend("topright", "Sloped baseline", bty = "n")
```

#### Peak filtering

The peak-finding algorithm may often detect a lot of peaks that are little more than noise. Thus, it is recommended to filter out extraneous peaks at this stage (especially if you are processing a lot of samples) as it can greatly reduce the computational burden of peak table construction. This can be accomplished directly by using the arguments `sd_max` (to filter by peak width) and/or `amp_thresh` (to filter by peak height). Alternatively, the `filter_peaks` function can be used to filter peaks after the `peak_list` has already been created.

### Peak table assembly

After obtaining a `peak_list`, the `get_peaks` function performs complete-linkage hierarchical clustering to link peaks across samples. It returns a `peak_table` object with samples as rows and peaks as columns. The `peak_table` object also has slots for holding metadata about peaks, samples, and the parameters used in the analysis. If you have a lot of samples, this step can be quite computationally expensive. Thus, it is suggested to filter the peak_list provided to `get_peaktable` in order to remove extraneous peaks (see *Peak finding and fitting* section above). Peaks can also be filtered after `peak_table` assembly using the `filter_peaktable` function. An important parameter here is `hmax` which controls the stringency with which retention times are matched across samples. A low value of `hmax` will increase the odds of splitting a single peaks across multiple columns, while a high value of `hmax` will increase the odds of erroneously combining multiple peaks into the same column.

```{r get-peaktable}
# assembly peak table from peak_list saved in `pks_egh`
pk_tab <- get_peaktable(pks_egh, response = "area", hmax = 0.2)

# print first six columns of peak table
head(pk_tab$tab[,1:6])
```

### Further analysis and data-visualization

#### Attaching metadata

To begin analyzing your peak table, you will usually want to attach sample metadata to the `peak_table` object. This can be easily accomplished using the `attach_metadata` function. This function takes a `metadata` argument which should be supplied with a `data.frame` containing experimental metadata, where one of the columns matches the names of your samples. This column should be specified by supplying the column name as a string to the `column` argument. This will attach the ordered metadata in the `sample_meta` slot of your peak table. The peak table can then be normalized (e.g., by dividing out the sample weight) using the `normalize_data` function.

```{r attach-metadata}
# load example metadata
path <- system.file("extdata", "Sa_metadata.csv", package = "chromatographR")
meta <- read.csv(path)
# attach metadata
pk_tab <- attach_metadata(peak_table = pk_tab, metadata = meta, column = "vial")
# normalize peak table by sample mass
pk_tab <- normalize_data(peak_table = pk_tab, column = "mass")
```

#### Attaching reference spectra

Optionally, you can attach scaled reference spectra to the peak_table using the `attach_ref_spectra` function. This can be helpful for working with UV spectra programmatically (e.g., to sort peaks by their chromophores). Reference spectra are defined either as the spectrum with the highest intensity for each peak (when `ref = "max.int"`) or as the spectrum with the highest average correlation to the other spectra associated with the peak (when `ref = "max.cor"`). Below, we show how these spectra can be used to construct a correlation matrix to find peaks matching a particular chromophore.

```{r attach-reference-spectra}
pk_tab <- attach_ref_spectra(pk_tab, ref = "max.int")
cor_matrix <- cor(pk_tab$ref_spectra)
hx <- names(which(cor_matrix[,"V20"] > .97))
matplot(x = as.numeric(rownames(pk_tab$ref_spectra)),
                   y = pk_tab$ref_spectra[, hx], type = 'l',
                   ylab = "Abs (mAU)", 
                   xlab = "Wavelength (nm)")
```

#### Data visualization

##### Mirror plot

The `mirror_plot` function provides a quick way to visually compare results across treatment groups.

```{r mirror-plot}
mirror_plot(pk_tab, lambdas = c(210), var = "trt", legend_size = 2)
```

##### Plotting spectra

The `plot_spectrum` function allows you to easily plot or record the spectra associated with a particular peak in your peak table. This is useful for interpreting your results and/or checking for errors in your peak table. For example, you may want to check if the spectra for a particular peak match across different samples, or you may want to compare your spectrum with a known standard. The `plot_spectrum` function can be used to plot only the spectrum or only the chromatographic trace using the arguments `plot_spectrum` and `plot_trace`. By default it will plot the trace and spectrum from the chromatogram with the largest peak in the peak table. Alternatively, you can choose the chromatogram index and wavelength using the `idx` and `lambda` arguments.

```{r plot-spectra, fig.height = 6, fig.alt = "Output of `plot_spectrum` function. Trace of chromatogram 4 at 318 nm (top) and spectrum of peak `V7` (bottom)."}
par(mfrow = c(2,1))
peak <- "V7"
plot_spectrum(peak, peak_table = pk_tab, chrom_list = warp_vpdtw,
              verbose = FALSE)
```

The `plot_spectrum` function can also be used to generate interactive plots using [plotly](https://plotly.com/r/).

```{r plotly-plot, eval=FALSE}
plot_spectrum(peak, peak_table = pk_tab, chrom_list = warp_vpdtw, 
              verbose = FALSE, engine = "plotly")
```

The `plot_all_spectra` function can be used to visually compare the spectra for a specified peak across all samples.

```{r plot-all-spectra, fig.alt="Output of `plot_all_spectra` function showing scaled spectra of peaks `V13` and `V14`."}
par(mfrow=c(1,2))
plot_all_spectra("V13", peak_table = pk_tab, export = FALSE)
plot_all_spectra("V14", peak_table = pk_tab, export = FALSE)
```

Inconsistencies in the spectra (as can be seen with "V13") can be indicative of a problem with the alignment or co-elution of multiple substances during the same retention window. By contrast, "V14" looks relatively pure.

The new `plot_spectrum_inset` function can also be used to plot spectra onto the chromatogram.

```{r inset-plot, eval=requireNamespace("ggplot2", quietly=TRUE), fig.width=8, fig.height=6}
plot_spectrum_inset("V13", peak_table = pk_tab, 
              verbose = FALSE, xlab="Retention time (min)",
              position="top", inset_width = 0.25)
```

This function can also be used to produce multiple insets. The spectra can also be given titles using the `title` argument. Additional peaks can be labeled with arbitrary labels using the `annotate_peaks` function.

```{r double-inset-plot, eval=requireNamespace("ggplot2", quietly=TRUE), fig.width=8, fig.height=6}
plot_spectrum_inset(c("Unknown 1" = "V7",
                      "Unknown 2" = "V15"), idx=c(1:4), 
                    peak_table = pk_tab, 
              verbose = FALSE, xlab="Retention time (min)",
              position=list(c(0.0,0.6),c(0.55,0.4)), inset_width = 0.25,
              scale_spectrum = FALSE) |>
   annotate_peaks(loc = c(A = "V5", B = "V7", C="V17",
                          D = "V21", E = "V22"), peak_table = pk_tab)
```

##### Plot peak table function

The `plot.peak_table` function provides a simplified interface to various options for plotting data from the `peak_table`. For example, it can be used as a quick interface to the `plot_spectrum` and `plot_all_spectra` functions shown above. It can also be used to quickly compare results across treatments by calling `boxplot`.

```{r plot-peaktable, fig.height = 6}
par(mfrow = c(3,1))
plot(pk_tab, loc = "V13", box_plot = TRUE, vars = "trt", verbose = FALSE)
```

## References

::: {#refs}
:::

## Session Information

```{r display-session-info}
sessionInfo()
```
