---
title: "Frequently asked questions"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(chromatographR)
```

##### 1. How should I format my data for analysis in `chromatographR`?

`chromatographR` functions expect chromatograms to be in wide format with retention times as rownames. Thus, if you have a csv file with retention times in the first column, you should be able to read the files into R as follows:

```{r, eval = FALSE}
lapply(vector_of_file_paths, function(f){
  read.csv(f, row.names = 1)
})
```

If you have chromatograms in "long" format, you will need to reshape them into wide format before proceeding with the analysis.

\n

##### 2. Some of my peaks aren't being integrated when I run `get_peaks`.

Try adjusting the value of the `sd.max` argument, especially if you are seeing that your wider peaks are the ones that aren't being integrated. This argument controls the maximum peak width. We have tried to set the default value (`sd.max = 50`) to a sensible number, but what is "sensible" can vary with the resolution of your data. For example, in the example shown below, wider peaks will not be picked up with `sd.max = 2.5` but if we increase `sd.max` to `10`, we are able to successfully integrate all of our peaks. If this fails, try adjusting the `smooth_window`.

```{r}
data(Sa_pr)
par(mfrow=c(2,1))
pks <- get_peaks(Sa_warp, lambdas = "210", fit = "egh", sd.max = 2.5)
plot(pks, main = "'sd.max' = 3")

pks <- get_peaks(Sa_warp, lambdas = "210", fit = "egh", sd.max = 10)
plot(pks, main = "'sd.max' = 10")
```
